\documentclass[14pt]{beamer}

\usepackage[british]{babel}

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{color}

\usepackage{listings}
\usepackage{prettylistings}
\usepackage{lang-llvm}
\usepackage{lang-spl}
\usepackage{lang-ssm}

\usepackage{xspace}
\usepackage{url}

\usepackage{array}

\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\newcommand{\splang}{\textsc{Splang}\xspace}
\newcommand{\llvm}{\textsc{LLVM}\xspace}
\newcommand{\demo}{\raisebox{0.2cm}{\tiny[\textsc{Demo}]}\xspace}

\title{\splang}
\author{Wouter Geraedts \and Joshua Moerman}
\institute{Radboud Universiteit Nijmegen}
\date{}

\begin{document}

\begin{frame}
	\titlepage
\end{frame}

\begin{frame}
	Designing IR
	\bigskip

	\begin{itemize}
		\item For P4 we want to target \llvm.
		\item Thus IR should be general enough.
		\item \llvm is
			\begin{itemize}
				\item in between C and ASM
				\item Strongly typed (but no polymorphism)
			\end{itemize}
	\end{itemize}

	\lstinputlisting[language=llvm]{llvm-example.ll}
\end{frame}

\begin{frame}
	Two ways to deal with polymorphism
	
	\begin{itemize}
		\item Boxing
		\item Templating
	\end{itemize}
\end{frame}

\begin{frame}
	Boxing
	\bigskip
	
	\lstinputlisting[language=c]{boxing.c}
	+ Runtime casts (= reinterpret bits)
\end{frame}

\begin{frame}
	Templating
	\bigskip
	
	\lstinputlisting[language=spl]{templating.spl}
	Generates:
	\begin{itemize}
		\item $\mathsf{id}_{\mathtt{Int}}$
		\item $\mathsf{id}_{\mathtt{Bool}}$
		\item $\mathsf{id}_{[\mathtt{Int}]}$
		\item $\mathsf{id}_{(\mathtt{Int}, \mathtt{Int})}$
		\item ...
	\end{itemize}
	
	for any type needed.
\end{frame}

\begin{frame}
	We chose for templating.
	\bigskip
	
	Disadvantages
	\begin{itemize}
		\item \small Not every type-checked program compiles in finite time.
		\item \small More work implementationwise.
	\end{itemize}
	Advantages
	\begin{itemize}
		\item \small At each point the exact concrete type is known.
		\item \small We can store tuples on the stack instead of on the heap.
		\item \small No need for casts.
	\end{itemize}
\end{frame}

\begin{frame}
	Example: tuples in SSM
	
	\begin{columns}[T]
		\begin{column}{0.6\textwidth}
			\only<1>{\lstinputlisting[language=spl,firstline=1,lastline=1]{tuple.spl}}
			\only<2>{\lstinputlisting[language=spl,firstline=2,lastline=2]{tuple.spl}}
			\only<3>{\lstinputlisting[language=spl,firstline=3,lastline=3]{tuple.spl}}
		
			Becomes:
			\only<1>{\lstinputlisting[language=ssm,firstline=1,lastline=4]{tuple.ssm}}
			\only<2>{\lstinputlisting[language=ssm,firstline=5,lastline=6]{tuple.ssm}}
			\only<3>{\lstinputlisting[language=ssm,firstline=7,lastline=9]{tuple.ssm}}
		\end{column}
		\begin{column}{0.4\textwidth}
			\begin{tabular}{|C{4em}|l}
				\vdots			& \\ \cline{1-1}
				\only<1-2>{1}\only<3>{2}
								& \only<2>{$\leftarrow$ SP} \\ \cline{1-1}
				\only<1-2>{2}\only<3>{3}
								& \only<3>{$\leftarrow$ SP} \\ \cline{1-1}
				3				& \only<1>{$\leftarrow$ SP} \\ \cline{1-1}
								& \\
			\end{tabular}
			
			\vspace{1em}
		\end{column}
	\end{columns}	
\end{frame}

\begin{frame}
	\begin{center}
		Also check out:
		\url{https://github.com/Wassasin/splang}
	\end{center}
\end{frame}


\begin{frame}
	\begin{center}
		\Huge Questions?
	\end{center}
\end{frame}

\end{document}
